---
title: "fastymd"
output:
    html:
        meta:
            css: ["@default@1.13.67", "@callout@1.13.67", "@article@1.13.67"]
            js: ["@sidenotes@1.13.67", "@copy-button@1.13.67", "@callout@1.13.67", "@toc-highlight@1.13.67"]
        options:
            toc: true
            js_highlight:
                package: prism
                version: 1.29.0
        
vignette: >
  %\VignetteEngine{litedown::vignette}
  %\VignetteIndexEntry{fastymd}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{microbenchmark, fasttime, lubridate, ymd}
---

```{r, include = FALSE}
litedown::reactor(print = NA)
```

## Overview

fastymd is a package for working with Year-Month-Day (YMD) style date objects.
It provides extremely fast passing of character strings and numeric values to
date objects as well as fast decomposition of these in to their year, month
and day components. The underlying algorithms follow the
[approach of Howard Hinnant](https://howardhinnant.github.io/date_algorithms)
for calculating days from the [UNIX Epoch](https://en.wikipedia.org/wiki/Unix_time)
of [Gregorian Calendar](https://en.wikipedia.org/wiki/Gregorian_calendar) dates
and vice versa.

The API won't give any surprises:

```{r}
library(fastymd)
cdate <- c("2025-04-16", "2025-04-17")
(res <- fymd(cdate))
res == as.Date(cdate)
get_ymd(res)
fymd(2025, 4, 16) == res[1L]
```

Invalid dates will return `NA` and a warning:

```{r}
fymd(2021, 02, 29) # not a leap year
```

More interesting is the handling of output after a valid date. Consider the
following timestamp:

```{r}
timelt <- as.POSIXlt(Sys.time(), tz = "UTC")
(timestamp <- strftime(timelt, "%Y-%m-%dT%H:%M:%S%z"))
```

By default the time element is ignored:

```{r}
(res <- fymd(timestamp))
res == as.Date(timestamp, tz = "UTC")
```

This ignoring of the timestamp is both good and bad. For timestamps it makes
perfect sense, but perhaps you have simple dates and a concern that some are
corrupted. For these we can use the `strict` argument:

```{r}
cdate <- "2025-04-16nonsense "
fymd(cdate)
fymd(cdate, strict = TRUE)
```

## Benchmarks

::: {.callout-important data-legend="Comparison with fasttime::fastDate()"}

The character method  of `fymd()` parses input strings in a fixed, year, month
and day order. These values must be digits but can be separated by any non-digit
character. This is similar in spirit to the `fastDate()` function in Simon
Urbanek's [fasttime](https://CRAN.R-project.org/package=fasttime) package, using
pure text parsing and no system calls for maximum speed.

For extremely fast passing of POSIX style timestamps you will struggle to beat
the performance of [fasttime](https://CRAN.R-project.org/package=fasttime). This
works fantastically for timestamps that do not need validation and are within
the date range supported by the package (currently 1970-01-01 through to the
year 2199).

`fymd()` fills the, admittedly small, niche where you want fast parsing of YMD
strings along with date validation and support for a wider range of dates from
the [Proleptic Gregorian calendar](https://en.wikipedia.org/wiki/Proleptic_Gregorian_calendar)
(currently we support years in the range `[-9999, 9999]`). This additional
capability does come with a small performance penalty but, hopefully, this has
been kept to a minimum and the implementation remains competitive.

:::

```{r}
library(microbenchmark)

# 1970-01-01 (UNIX epoch) to "2199-01-01"
dates <- seq.Date(from = .Date(0), to = fymd("2199-01-01"), by = "day")

# comparison timings for fymd (character method)
cdates  <- format(dates)
(res_c <- microbenchmark(
    fasttime  = fasttime::fastDate(cdates),
    fastymd   = fymd(cdates),
    ymd       = ymd::ymd(cdates),
    lubridate = lubridate::ymd(cdates),
    check     = "equal"
))

# comparison timings for fymd (numeric method)
ymd  <- get_ymd(dates)
(res_n <- microbenchmark(
    fastymd   = fymd(ymd[[1]], ymd[[2]], ymd[[3]]),
    lubridate = lubridate::make_date(ymd[[1]], ymd[[2]], ymd[[3]]),
    check     = "equal"
))

# comparison timings for year getter
(res_get_year <- microbenchmark(
    fastymd   = get_year(dates),
    ymd       = ymd::year(dates),
    lubridate = lubridate::year(dates),
    check     = "equal"
))

# comparison timings for month getter
(res_get_month <- microbenchmark(
    fastymd   = get_month(dates),
    ymd       = ymd::month(dates),
    lubridate = lubridate::month(dates),
    check     = "equal"
))

# comparison timings for mday getter
(res_get_mday <- microbenchmark(
    fastymd   = get_mday(dates),
    ymd       = ymd::mday(dates),
    lubridate = lubridate::day(dates),
    check     = "equal"
))
```
